---
title: "Data Cleaning & Preprocessing"
author: "530318646, "
date: today
repository: https://github.com/DipsyCdua/reef.git
format: 
  html:
    embed-resources: true
    code-fold: true
    code-tools: true
    theme: sandstone
    fig_caption: yes
    table-of-contents: true
    toc: true
    toc_depth: 4
    toc_float: true
execute:
  echo: true
  tidy: true
number-sections: false
---
```{r setup, echo=FALSE}
library(tidyverse)
library(ncdf4)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(readr)
library(lubridate)
library(sf)
library(fs)
library(viridis)
library(terra)
library(mapview)
library(vroom)

# database libraries
library(DBI)
library(RSQLite)
library(RPostgres)

set.seed(3926)

# in order to not exceed the github repository storage, all the raw data we downloaded from our sources are in zipped folders

# unzip data folders
zip_folder <- "raw_input_data"

zip_files <- list.files(path = zip_folder, pattern = "\\.zip$", full.names = TRUE)

for (zip_file in zip_files) {
  output_dir <- tools::file_path_sans_ext(zip_file)
  
  # only unzip if the output directory doesn't already exist
  if (!dir.exists(output_dir)) {
    unzip(zip_file, exdir = output_dir)
    message("Extracted: ", zip_file)
  } else {
    message("Already extracted: ", zip_file)
  }
}
```

# Disclaimer!

This .qmd file contains all of the data cleaning, pre-processing and output for our project. We included this as a separate file to the Shiny App and modelling files because of how long it takes to run this code. So, **this code has already been run and the cleaned data output is already saved in this github (in output_data)**. If you were to run the code, it would simply over-write the existing cleaned data without making any changes.


# Datasets

We have downloaded and used the following datasets in our project:

#### 1. NOAA Coral Reef Watch Satellite Coral Bleaching Heat Stress Products

**Temperature variables**: Degree-heating weeks (DHW), sea-surface temperature (SST) and sea-surface temperature anomalies (SSTA)
**Date**: 1985-2025

3 folders, 1 for each temperature variable, was provided to us. Each folder contained numerous NetCDF files with temperature data of the entire GBR region. Each NetCDF file contained lon, lat and temperature data for each month of each year from 1985-2025.

#### 2. NOAA El Niño Southern Oscillation Index data

**SOI Index**: SOI index anomalies (what we wanted to use) and SOI standardized index.

**Date:** 1951-2025

This data was downloaded as a single text file and cleaned in Python (`"soi_cleaning.py"`), then downloaded as a clean csv file in the output_data folder. It was also merged into our final, cleaned csv file.


#### 3. Geoscience Australia GBR Bathymetry Data

This folder contained a TIFF file with bathymetry data of the entire GBR (ocean depth). We used this ocean depth (metres) to categories regions into continental shelf categories; inner shelf [0, 20), mid shelf [20, 40) and outer shelf [40, inf).

We then classified points on the GBR with a shelf value. We als merged regions to create 3 multipolygons that were downloaded as a Shape file for our Shiny App map visualisation.

#### 4. GBR Marine Park Authority Boundary

These shape files will be used to filter out geographic points to only include coordinates that lie within the official GBR marine park boundary.

# Data Cleaning Pipeline

## NOAA Temperature Data

### From Daily NetCDF to Monthly CSV

1. Extract longitude, latitude and temperature variable into dataframe

2. Pivot from wide to long format

3. Delete "land" coordinates

4. Calculate average monthly temperature for every point in the GBR

5. Create date column using file titles

6. Only keep ocean data within the GBR Marine Park zone (using shape file)

7. Combine different temperature csv files by month, then year, then join all years into 1 big csv

8. Drop years with incomplete data (now 1987-2024)

9. Using bathymetry data, create continental shelf zones (multipolygons) based on depth and assign shelf id to each point based on spatial join

10. Sreate shelf shape file for shiny app

11. Left Join the data with monthly SOI (all points in same month have same SOI)

12. Download as csv for ease of modelling & shiny

Due to sheer size of data, we decided to sacrifice the temporal resolution for high spatial resolution. We now have average daily temperature for each month from 1985-2025 for every single point.

We now have 3 folders for `mean_dhw`, `mean_sst` and `mean_ssta`.

5. Save as new csv file


#### DHW

```{r dhw-csv}
# file directories
input_dir <- "raw_input_data/Reef3/DHW"
output_dir <- "raw_input_data/Reef3/DHW/summarised"
dir_create(output_dir)

# nc files
nc_files <- dir_ls(input_dir, regexp = "\\.nc$")

# for loop to convert nc files to csvs
for (file_path in nc_files) {
  file_name <- basename(file_path)

  # if file is empty, ignore
  if (file_info(file_path)$size == 0) {
    cat("Skipping (empty file):", file_name, "\n")
    next
  }
  
  # open nc file and check if corrupted
  nc <- try(nc_open(file_path), silent = TRUE)
  if (inherits(nc, "try-error")) {
    cat("Skipping (error opening):", file_name, "\n")
    next
  }
  
  # variables
  lon <- ncvar_get(nc, "lon")
  lat <- ncvar_get(nc, "lat")
  time <- ncvar_get(nc, "time")
  dhw <- ncvar_get(nc, "degree_heating_week")
  nc_close(nc)

  # var dims
  n_lon <- length(lon)
  n_lat <- length(lat)
  n_time <- length(time)

  # wide to long format
  df <- expand.grid(
    lon = as.numeric(lon),
    lat = as.numeric(lat),
    time = as.numeric(time),
    KEEP.OUT.ATTRS = FALSE,
    stringsAsFactors = FALSE
  ) |>
    arrange(time, desc(lat), lon) |>
    mutate(dhw = as.vector(dhw)) |>
    filter(!is.na(dhw)) # dropping all "land points" for storage (where dhw=NaN)

  # summarise df + calculate monthly mean DHW for each location
  summary_df <- df |>
    group_by(lon, lat) |>
    summarise(
      mean_dhw = mean(dhw, na.rm = TRUE),
      .groups = "drop"
    )

  # save as csv
  file_name <- tools::file_path_sans_ext(basename(file_path))
  write.csv(summary_df, file.path(output_dir, paste0(file_name, "_summary.csv")))
  
  cat("Saved:", file_name, "\n")
}


# ----TEST----
# take a look at a random csv
# all of them have same dim (61,842 by 3) and same points
test = read.csv("raw_input_data/Reef3/DHW/summarised/combined_2023_03_GBR_DHW_summary.csv")
```


#### SST

```{r sst-csv}
# file directories
input_dir <- "raw_input_data/Reef3/SST"
output_dir <- "raw_input_data/Reef3/SST/summarised"
dir_create(output_dir)

# nc files
nc_files <- dir_ls(input_dir, regexp = "\\.nc$")

# for loop to convert nc files to csvs
for (file_path in nc_files) {
  file_name <- basename(file_path)
  
  # if file is empty, ignore
  if (file_info(file_path)$size == 0) {
    cat("Skipping (empty file):", file_name, "\n")
    next
  }
  
  # open nc file and check if corrupted
  nc <- try(nc_open(file_path), silent = TRUE)
  if (inherits(nc, "try-error")) {
    cat("Skipping (error opening):", file_name, "\n")
    next
  }
  
  # check if required variables exist
  has_var <- function(nc, varname) {
    varname %in% names(nc$var) || varname %in% names(nc$dim)
  }
  
  if (!all(sapply(c("lon", "lat", "time", "analysed_sst"), function(v) has_var(nc, v)))) {
    cat("Skipping (missing expected variables):", file_name, "\n")
    nc_close(nc)
    next
  }

  
  # variables
  lon <- ncvar_get(nc, "lon")
  lat <- ncvar_get(nc, "lat")
  time <- ncvar_get(nc, "time")
  sst <- ncvar_get(nc, "analysed_sst")
  nc_close(nc)

  # var dims
  n_lon <- length(lon)
  n_lat <- length(lat)
  n_time <- length(time)

  # wide to long format
  df <- expand.grid(
    lon = as.numeric(lon),
    lat = as.numeric(lat),
    time = as.numeric(time),
    KEEP.OUT.ATTRS = FALSE,
    stringsAsFactors = FALSE
  ) |>
    arrange(lon, lat, time) |>
    mutate(sst = as.vector(sst)) |>
    filter(!is.na(sst)) # dropping all "land points" for storage (where SST=NaN)

  # summarise df + calculate monthly mean SST for each location
  summary_df <- df |>
    group_by(lon, lat) |>
    summarise(
      mean_sst = mean(sst, na.rm = TRUE),
      .groups = "drop"
    )

  # save as csv
  file_name <- tools::file_path_sans_ext(basename(file_path))
  write.csv(summary_df, file.path(output_dir, paste0(file_name, "_summary.csv")))
  
  cat("Saved:", file_name, "\n")
}



# ----TEST----
# take a look at a random csv
# all of them have same dim (66,840 by 3) and same points
test <- read.csv("raw_input_data/Reef3/SST/summarised/combined_1990_08_GBR_SST_summary.csv")

fail_check <- nc_open("raw_input_data/Reef3/SST/combined_1986_05_GBR_SST.nc")
names(fail_check$var)
nc_close(fail_check)
```


#### SSTA

```{r ssta-csv}
# file directories
input_dir <- "raw_input_data/Reef3/SSTA"
output_dir <- "raw_input_data/Reef3/SSTA/summarised"
dir_create(output_dir)

# nc files
nc_files <- dir_ls(input_dir, regexp = "\\.nc$")

# for loop to convert nc files to csvs
for (file_path in nc_files) {
  file_name <- basename(file_path)

  # if file is empty, ignore
  if (file_info(file_path)$size == 0) {
    cat("Skipping (empty file):", file_name, "\n")
    next
  }
  
  # open nc file and check if corrupted
  nc <- try(nc_open(file_path), silent = TRUE)
  if (inherits(nc, "try-error")) {
    cat("Skipping (error opening):", file_name, "\n")
    next
  }
  
  # check if required variables exist
  has_var <- function(nc, varname) {
    varname %in% names(nc$var) || varname %in% names(nc$dim)
  }
  
  if (!all(sapply(c("lon", "lat", "time", "sea_surface_temperature_anomaly"), function(v) has_var(nc, v)))) {
    cat("Skipping (missing expected variables):", file_name, "\n")
    nc_close(nc)
    next
  }

  
  # variables
  lon <- ncvar_get(nc, "lon")
  lat <- ncvar_get(nc, "lat")
  time <- ncvar_get(nc, "time")
  ssta <- ncvar_get(nc, "sea_surface_temperature_anomaly")
  nc_close(nc)

  # var dims
  n_lon <- length(lon)
  n_lat <- length(lat)
  n_time <- length(time)

  # wide to long format
  df <- expand.grid(
    lon = as.numeric(lon),
    lat = as.numeric(lat),
    time = as.numeric(time),
    KEEP.OUT.ATTRS = FALSE,
    stringsAsFactors = FALSE
  ) |>
    arrange(time, desc(lat), lon) |>
    mutate(ssta = as.vector(ssta)) |>
    filter(!is.na(ssta)) # dropping all "land points" for storage (where SSTA=NaN)

  # summarise df + calculate monthly mean SSTA for each location
  summary_df <- df |>
    group_by(lon, lat) |>
    summarise(
      mean_ssta = mean(ssta, na.rm = TRUE),
      .groups = "drop"
    )

  # save as csv
  file_name <- tools::file_path_sans_ext(basename(file_path))
  write.csv(summary_df, file.path(output_dir, paste0(file_name, "_summary.csv")))
  
  cat("Saved:", file_name, "\n")
}



# ----TEST----
# take a look at a random csv
# all of them have same dim (66,840 by 3) and same points
test <- read.csv("raw_input_data/Reef3/SSTA/summarised/combined_1985_01_GBR_SSTA_summary.csv")

fail_check <- nc_open("raw_input_data/Reef3/SSTA/combined_1986_05_GBR_SSTA.nc")
names(fail_check$var)
nc_close(fail_check)
```


### Combine monthly data in same years

```{r yearly-csvs}
#----YEARLY CSVS---
combine_monthly_files_by_year <- function(var, input_dir, output_dir) {
  # make output dir
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  # list all relevant files
  files <- list.files(input_dir, pattern = paste0(".*_", var, "_summary\\.csv$"), full.names = TRUE)

  # extract year and month from filename
  file_info <- tibble(
    file = files,
    year = str_extract(files, "(?<=combined_)\\d{4}") |> as.integer(),
    month = str_extract(files, "(?<=combined_\\d{4}_)\\d{2}") |> as.integer()
  )

  # group by year
  file_info_split <- split(file_info, file_info$year)

  walk(file_info_split, function(group) {
    year = unique(group$year)

    # read and tag each monthly file
    dfs <- map2(group$file, group$month, function(f, m) {
      df <- read.csv(f, show_col_types = FALSE) |>
        mutate(year = year, month = m)

      # if variable is DHW or SSTA, sort lat ascending to fix flipped map issue
      if (var %in% c("DHW", "SSTA")) {
        df <- df |> arrange(lat)
      }

      df
    })

    # combine into one data frame
    df_year <- bind_rows(dfs)

    # standardise column names
    var_col <- paste0("mean_", tolower(var))
    df_year <- df_year |> select(lon, lat, year, month, !!var_col)

    # write to CSV
    out_file <- file.path(output_dir, paste0("monthly_", year, "_", var, "_summary.csv"))
    write.csv(df_year, out_file)
    cat("Wrote:", out_file, "\n")
  })
}

# apply to data
combine_monthly_files_by_year("DHW", "raw_input_data/Reef3/DHW/summarised", "raw_input_data/Reef3/DHW/summarised/yearly")
combine_monthly_files_by_year("SST", "raw_input_data/Reef3/SST/summarised", "raw_input_data/Reef3/SST/summarised/yearly")
combine_monthly_files_by_year("SSTA", "raw_input_data/Reef3/SSTA/summarised", "raw_input_data/Reef3/SSTA/summarised/yearly")

```


### Yearly CSV File for all variables
```{r combine-vars}
# combine all 3 variables
merge_yearly_temp_data <- function(years = 1985:2025,
                                   dhw_dir,
                                   sst_dir,
                                   ssta_dir,
                                   base_output_dir) {

  # create "combined" folder if it doesn't exist
  output_dir <- file.path(base_output_dir, "combined")
  if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
  
  # find the corresponding, dhw, sst and ssta file for that year
  for (yr in years) {
    dhw_file <- file.path(dhw_dir, paste0("monthly_", yr, "_DHW_summary.csv"))
    sst_file <- file.path(sst_dir, paste0("monthly_", yr, "_SST_summary.csv"))
    ssta_file <- file.path(ssta_dir, paste0("monthly_", yr, "_SSTA_summary.csv"))

                     
    # helper to read csv or return NULL
    safe_read <- function(filepath, var) {
      if (file.exists(filepath)) {
        read.csv(filepath, show_col_types = FALSE) |>
        rename(!!paste0("mean_", tolower(var)) := matches(paste0("mean_", tolower(var))))
    } else {
    NULL
      }
      }

    # read in var data
    dhw <- safe_read(dhw_file, "DHW")
    sst <- safe_read(sst_file, "SST")
    ssta <- safe_read(ssta_file, "SSTA")


    # skip year if all are NULL or empty
    if (all(map_lgl(list(dhw, sst, ssta), ~ is.null(.) || nrow(.) == 0))) {
      message("Skipping year ", yr, " (no data found)")
      next
    }

    # merge
    combined <- list(dhw, sst, ssta) |>
      compact() |>  # remove nans
      reduce(full_join, by = c("lon", "lat", "year", "month")) |>
      arrange(month, lat, lon)
    
    # drop rows where any column is na
    combined <- drop_na(combined)
    
    # round all vars to 2 decimal places
    combined <- dplyr::mutate(combined, dplyr::across(c(mean_dhw, mean_sst, mean_ssta), ~ round(., 2)))
    
    # write merged file
    output_path <- file.path(output_dir, paste0("combined_", yr, "_GBR_allvars.csv"))
    write.csv(combined, output_path)
    message("Saved merged file for year: ", yr)
  }

  message("All available yearly data merged and saved to: ", output_dir)
}



# merge data
merge_yearly_temp_data(
  dhw_dir = "raw_input_data/Reef3/DHW/summarised/yearly",
  sst_dir = "raw_input_data/Reef3/SST/summarised/yearly",
  ssta_dir = "raw_input_data/Reef3/SSTA/summarised/yearly",
  base_output_dir = "raw_input_data/Reef3"
)
```

### Fix Data Types

Create date column and geospatial point column.

```{r}
# noaa data
noaa_dir <- "raw_input_data/Reef3/combined"

# output subfolder
output_dir <- file.path(noaa_dir, "with_date")
if (!dir.exists(output_dir)) dir.create(output_dir)

# lisy of all files
csv_files <- list.files(noaa_dir, pattern = "\\.csv$", full.names = TRUE)

# fix columns
walk(csv_files, function(file) {
  df <- read.csv(file, show_col_types = FALSE)

  # Convert to proper Date object (first day of each month)
  df <- df |> 
    mutate(date = ymd(sprintf("%04d-%02d-01", year, month))) |> # yyyy-mm-01
    select(-year, -month) |>
    st_as_sf(coords = c("lon", "lat"), crs = 4326) |>  # EPSG:4326 = WGS 84
    mutate(geometry = st_as_text(geometry))  # store as WKT string

  # Save to new subfolder
  out_file <- file.path(output_dir, basename(file))
  write.csv(df, out_file)
  cat("Saved:", out_file, "\n")
})
```

### Drop points outside GBR

We will drop all data outside of the official GBR Marine Park region

```{r}
input_dir <- "raw_input_data/Reef3/combined/with_date"
output_dir <- file.path(input_dir, "within_gbr")
dir_create(output_dir)

# gbr polygon and transform to wkt
gbr_area <- st_read("raw_input_data/Great_Barrier_Reef_Marine_Park_Boundary_94_-2154764915499037387/Great_Barrier_Reef_Marine_Park_Boundary.shp")
gbr_area <- st_transform(gbr_area, crs = 4326)

# helper function to filter and save new data from each csv
filter_and_save <- function(file_path) {
  message("Processing: ", basename(file_path))
  
  # read csv and make sf
  df <- read.csv(file_path)
  sf_df <- st_as_sf(df, wkt = "geometry", crs = 4326)
  
  # filter points in gbr_area
  inside <- st_within(sf_df, gbr_area, sparse = FALSE)[,1]
  sf_filtered <- sf_df[inside, ]
  
  # write into output folder
  output_path <- file.path(output_dir, basename(file_path))
  write.csv(sf_filtered, output_path)
}

# now apply function to all files
csv_files <- dir_ls(input_dir, glob = "*.csv")
walk(csv_files, filter_and_save)

# CHECK

check_data <- read.csv("raw_input_data/Reef3/combined/with_date/within_gbr/combined_1985_GBR_allvars.csv")

check_data <- check_data |>
  mutate(
    lon = as.numeric(str_extract(geometry, "(?<=c\\().+?(?=,)")),
    lat = as.numeric(str_extract(geometry, "(?<=, ).+?(?=\\))"))
  )

ggplot(check_data, aes(x = lon, y = lat, fill = mean_dhw)) +
  geom_tile() +
  scale_fill_viridis_c(option = "C", na.value = "grey90") +
  coord_fixed() +
  labs(
    title = "Mean DHW",
    fill = "Mean DHW"
  ) +
  theme_minimal()
```


### Combining all CSV files into 1 dataframe

```{r}
# PLEASE NOTE I PERFORMED THIS CODE ONCE WITH DIFFERENT PATHS, SO NOW THESE CSVS ARE IN A NEW FOLDER "noaa_data" WHEN YOU WILL FIND "Reef3/combined/with_date/within_gbr"

# SO FROM NOW ON, REEF CSVS WILL BE IN "noaa_data" which you have

# for group members: this will be the directory to the "NOAA_data" folder in the Google drive

#----REEF DATA----
data_dir <- "raw_input_data/noaa_data"
csv_files <- list.files(data_dir, pattern = "\\.csv$", full.names = TRUE)

# read and format each csv
load_and_process_csv <- function(file) {
  df <- read.csv(file)
  df <- df|>
    mutate(
      year = year(as.Date(date)),
      # extract lon and lat from "c(lon, lat)"
      lon = as.numeric(str_extract(geometry, "(?<=c\\()[^,]+")),
      lat = as.numeric(str_extract(geometry, "(?<=,\\s)-?\\d+\\.\\d+(?=\\))"))
    )
  return(df)
}

# combine all csv files to 1 dataframe
all_data <- bind_rows(lapply(csv_files, load_and_process_csv)) 

# drop old geom
all_data <- all_data |>
  select(-geometry)
```


### Only include years with full data

Some years only contained a few months of data. Because we know month is a huge factor for temperature (e.g. summer months have greater temperatures), we only included years with 12 months of data.

New year range: 1987-2024
```{r}
# find years with less than 12 months (1985, 1986, 2025)
incomplete_years <- all_data |>
  mutate(year = year(date),
         month = month(date)) |>
  distinct(year, month) |>
  count(year) |>
  filter(n < 12) |>
  pull(year)

# drop these years
all_data_clean <- all_data |>
  filter(!year %in% incomplete_years) |>
  select(-year)
  
# check
incomplete_years_clean <- all_data_clean |>
  mutate(year = year(date),
         month = month(date)) |>
  distinct(year, month) |>
  count(year) |>
  filter(n < 12) |>
  pull(year)

```

## Bathymetry Data

### Creating new "Shelf" boundaries
```{r}
bathy <- rast("raw_input_data/Great Barrier Reef Bathymetry 2020 100m/Great_Barrier_Reef_2020_100m_MSL_cog.tif")

# look (wow)
plot(bathy, main = "Great Barrier Reef Bathymetry (2020, 100m)")

# reduce to only gbr
gbr_vect <- vect(gbr_area)
bathy_crop <- crop(bathy, gbr_vect)
bathy_gbr <- mask(bathy_crop, gbr_vect)

plot(bathy_gbr, main = "GBR Bathymetry (Clipped)")

# create new shelf variable based on bathymetry data
bathy_depth <- abs(bathy_gbr)

shelf_matrix <- matrix(c(
  0, 20, 1,  # inner shelf (1)
  20, 40, 2, # mid shelf (2)
  40, Inf, 3 # outer shelf (3)
), ncol = 3, byrow = TRUE)

# classify!
shelf_class <- classify(bathy_depth, shelf_matrix)
shelf_class <- subst(shelf_class, c(1, 2, 3), c("I", "M", "O")) # rename

# see new shelf
plot(shelf_class, col = c("skyblue", "dodgerblue", "navy"), main = "GBR Shelf Zones")
```

### Create shape file for Shiny
```{r}
# MAKE POLYGONS
# lets reduce resolution by factor of 10 cause its too high res
shelf_class_down <- terra::aggregate(shelf_class, fact = 10, fun = modal, na.rm = TRUE)

# try now
shelf_poly_raw <- terra::as.polygons(shelf_class_down, dissolve = FALSE)
names(shelf_poly_raw) <- "shelf_zone"

# dissolve polygons by shelf
shelf_poly <- terra::aggregate(shelf_poly_raw, by = "shelf_zone", fun = mean)

# convert to sf
shelf_sf <- st_as_sf(shelf_poly) |>
  mutate(
    shelf_class = factor(shelf_zone, levels = c(1, 2, 3), labels = c("Inner", "Mid", "Outer"))
  )


# join the close multipolygons together if they are touching and same class
shelf_sf_dissolved <- shelf_sf |>
  group_by(shelf_class) |>
  summarise(geometry = st_union(geometry), .groups = "drop")

# look at shelf
ggplot(shelf_sf_dissolved) +
  geom_sf(aes(fill = shelf_class), colour = NA) +
  coord_sf()

# look (yes looks good)
ggplot(shelf_sf) +
  geom_sf(aes(fill = shelf_class), colour = NA) +
  scale_fill_manual(
    values = c("Inner" = "#1b9e77", "Mid" = "#d95f02", "Outer" = "#7570b3"),
    name = "Shelf Zone"
  ) +
  coord_sf()


# SAVE
# make a folder
out_dir <- "output_data/shelf_boundaries"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
# save as a shape file for map overlay
out_path <- file.path(out_dir, "shelf_zones_dissolved.shp")
st_write(shelf_sf_dissolved, out_path, delete_dsn = TRUE)
```

### Add new column `shelf` using cleaned bathymetry data
```{r}
# make vector of points from all_data_clean (lon, lat)
points_vect <- vect(all_data_clean, geom = c("lon", "lat"), crs = crs(shelf_class))

# make sure it is same class as bathymetry data
points_vect <- project(points_vect, crs(shelf_class))

# extract shelf zone for each point from shelf_class
shelf_values <- terra::extract(shelf_class, points_vect)

# bind shelf values to data
all_data_clean$shelf <- shelf_values[, 2]  # column 2 = extracted class

# check some random year
points_1990 <- all_data_clean |>
  filter(date == as.Date("1990-01-01"))

ggplot(points_1990, aes(x = lon, y = lat, colour = shelf)) +
  geom_point(size = 0.5, alpha = 0.8) +
  scale_colour_manual(values = c("I" = "skyblue", "M" = "dodgerblue", "O" = "navy")) +
  coord_fixed() +
  labs(title = "GBR Shelf Classification on 1990-01-01",
       colour = "Shelf Zone") +
  theme_minimal()
```

### Join Data with SOI Index & Download (NOW COMPLETE)
```{r}
# make sure soi has same "date"
soi <- dplyr::rename(soi, date = Date)

# final join
reef03_data <- all_data_clean |>
  left_join(soi, by = "date")

# drop year 2025!!!
reef03_data_clean <- reef03_data |>
  filter(date <= "2024-12-01")

# download data
write.csv(reef03_data, file = "output_data/NEW_reef_full.csv")

# load in downloaded data when needed
reef03_download <- vroom(
  file      = "output_data/NEW_reef_full.csv",
  delim     = ",",
  col_types = cols(date = col_date(format = ""))  # parse date up front
)
```

# Database

Initially, we thought it would be a good idea to put all of our data into an SQL database. We believed it would be easier for all the group members to access the data and potentially improve the performance of our Shiny App. However, after consulting with our tutor, this turned out to be redundant and it was indeed easier to just use data from a static csv file (which is what we ended up doing).

Therefore, this code is not relevant to our project. However, it demonstrates evidence of how we tried to troubleshoot the data scalability issue.

## SQL schema

::: panel-tabset

### Create Schema (DON'T RUN)
```{r con}
# connection
con <- dbConnect(
  RPostgres::Postgres(),
  host = "localhost",
  dbname = "reef03_noaa_db",
  user = "postgres",
  password = "123imanepostgres",
  port = 5432
)

# check if empty
dbListTables(con)
```

```{r schema}
# postgis
dbExecute(con, "
CREATE EXTENSION IF NOT EXISTS postgis;
")


# soi_data
dbExecute(con, "
DROP TABLE IF EXISTS soi_data;")

dbExecute(con, "
CREATE TABLE soi_data (
    soi_id SERIAL PRIMARY KEY,
    date DATE,
    soi_standard DOUBLE PRECISION,
    soi_anomaly DOUBLE PRECISION
);
")

# regions
# dbExecute(con, "
# DROP TABLE IF EXISTS regions;")
# 
# dbExecute(con, "
# CREATE TABLE regions (
#     region_id SERIAL PRIMARY KEY,
#     region_name VARCHAR(50),
#     geom GEOMETRY(Polygon, 4326)
# );
# ")


# reef_locations
dbExecute(con, "
DROP TABLE IF EXISTS reef_locations;")

dbExecute(con, "CREATE TABLE reef_locations (
    location_id SERIAL PRIMARY KEY,
    geom GEOMETRY(Point, 4326)
);
")

# environmental_data
dbExecute(con, "
DROP TABLE IF EXISTS environmental_data;")

dbExecute(con, "
CREATE TABLE environmental_data (
    data_id SERIAL PRIMARY KEY,
    location_id INT REFERENCES reef_locations(location_id),  -- foreign key
    date DATE,
    mean_sst DOUBLE PRECISION,
    mean_dhw DOUBLE PRECISION,
    mean_ssta DOUBLE PRECISION,
    FOREIGN KEY (location_id) REFERENCES reef_locations(location_id) ON DELETE CASCADE
);
")

# indexes
dbExecute(con, "CREATE INDEX idx_environmental_data_date ON environmental_data (date)")
dbExecute(con, "CREATE INDEX idx_soi_data_date ON soi_data (date)")
dbExecute(con, "CREATE INDEX idx_reef_locations ON reef_locations (location_id)")

# see
dbListTables(con)
```

::: .tabset

#### soi_data (DON'T RUN)
```{r soi-db}
# soi data
dbExecute(con, "DELETE FROM soi_data")
dbWriteTable(con, "soi_data", soi, append = TRUE, row.names = FALSE)
dbGetQuery(con, "SELECT * FROM soi_data;")
```

#### reef_locations

```{r reef-loc-db}
# forgot to add constraint
dbExecute(con, "
ALTER TABLE reef_locations
ADD CONSTRAINT unique_geom UNIQUE (geom);")

# just look at structure
loc_data <- read.csv("/Users/imanelattab/Desktop/DATA3888/reef_dbs/noaa_data/with_date/combined_2024_GBR_allvars.csv")

# extract unique spatial points (to avoid duplicates)
unique_points <- loc_data |> 
  select(geometry) |>
  distinct()

# insert unique spatial points into the reef_locations table
for (i in 1:nrow(unique_points)) {
  query = sprintf("
    INSERT INTO reef_locations (geom)
    VALUES (ST_GeomFromText('%s', 4326))
    ON CONFLICT (geom) DO NOTHING; 
  ", unique_points$geometry[i])
  dbExecute(con, query)
}

# check
dbGetQuery(con, "SELECT * FROM reef_locations LIMIT 5;")
```

#### environmental_data

```{r env-db}
# again forgot constraint
dbExecute(con, "
ALTER TABLE environmental_data
ADD CONSTRAINT fk_location_id FOREIGN KEY (location_id)
REFERENCES reef_locations(location_id)
ON DELETE CASCADE;")

# function for inserting environmental data from a single file
insert_environmental_data <- function(file) {
  data = read.csv(file)
  
  for (i in 1:nrow(data)) {
    # find  location_id based on the geometry
    query_location_id <- sprintf("
      SELECT location_id 
      FROM reef_locations 
      WHERE geom = ST_GeomFromText('%s', 4326);
    ", data$geometry[i])
    
    # get location_id
    location_id <- dbGetQuery(con, query_location_id)$location_id
    
    # insert environmental data into environmental_data table
    query_env_data <- sprintf("
      INSERT INTO environmental_data (location_id, date, mean_sst, mean_dhw, mean_ssta)
      VALUES (%d, '%s', %f, %f, %f);
    ", location_id, data$date[i], data$mean_sst[i], data$mean_dhw[i], data$mean_ssta[i])
    
    dbExecute(con, query_env_data)
  }
}

# get the list of all files to process
files <- list.files(path = "/Users/imanelattab/Desktop/DATA3888/reef_dbs/noaa_data/with_date", full.names = TRUE)

# loop through each file and insert the data
for (file in files) {
  insert_environmental_data(file)
}

```

:::


# Close Connection
```{r}
dbDisconnect(sqlite_con)
dbDisconnect(con)
```